Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-05-07 SIR_manuscript.tex
changes in methods

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where the term $\frac{\beta}{\gamma}$ is the classical basic reproduction number for a SIR model without testing and isolation (see, e.g., \cite{keeling2011modeling}), and $\Delta$ is the reduction parameter defined as follows. \ali{$\Delta$ should be updated, the expression will be linear in $\theta\_c$ but quadratic in $\theta\_w$, C stay unchanged. It can be proved that $\Delta$ increases in $\theta\_w$ for feasible range of $\theta\_w$ (I think this argument will go to the appendix).?}
\begin{equation}
\label{eq:Delta1}
\Delta= (A~ \theta\_w + B~ \theta\_c) / C,
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
A=& \gamma \Big( \big(\gamma(\omega+\gamma) + \omega p_I \hat F_I \big) \frac{(1-\theta\_w)S\_n^*}{N_0} + (\omega+\gamma) \hat F_I \Big),\\
B=& \omega p_I(\gamma \frac{S\_u^*}{N_0}+\omega)\hat F_I, \\ 
C=& (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big).
\end{aligned}
\end{equation}
Further details of derivation of $\Rnum$ are provided in section \ref{app:R0} of the appendix. Note that $\Delta$ has been presented in a specific monotonic form with respect to isolation efficacy parameters, $\theta\_w$ and $\theta\_c$, in order to simplify showing one of the results of this paper which is the reduction effect of isolation on $\Rnum$. Further, we represent $\Delta$ in another equivalent form in the appendix (Sec. \ref{app:rho}) to show the nontrivial-nonmonotonic effect of testing intensity and test return rate on $\Rnum$.

@@

The expression of $\Rnum$ has a complicated form with all of the model parameters involved. This expression can be simplifed and represented given the specific form of matrix $G_{11}$ \eqref{G11}. For the purpose of simplicity we present $\Rnum$ in the manuscript in terms of expressions $A$, $B$ and $C$, specified in \eqref{eq:abc}. This is intended to show that $\Rnum$ decreases as the strength of isolation, shown by $\theta\_w$ and $\theta\_c$, increases.

@@

This is to provide a mathematical materials to prove that $\pro \Delta$ can be positive or negative, with $\Delta$ defined in Eq.\,\eqref{eq:abc}, and thus $\pro \Rnum < 0$, where $\Rnum$ is the basic reproduction number, given in Eq.\,\eqref{R0}. We can rewrite matrix $G_{11}$ in \eqref{G11} in the following form to simplify the calculations.
\begin{equation}
\label{G112}
G_{11} = \frac{\beta}{\gamma \, C} 
% \left[ \begin {array}{c}  (\omega-\rho)/\omega \\ (1-\theta\_w)\rho/\omega  \end {array} \right]
\left[ \begin {array}{c}  S\_u^*/N_0 \\ (1-\theta\_w)\,S\_n^*/N_0  \end {array} \right]
\left[ \begin {array}{cccc} 
C-C_1, C-C_1-C_2\end {array} \right],
\end{equation}
where $C$ is the same as the one in Eq. \eqref{eq:abc}, i.e., 
$$C=(\omega+\gamma)(\gamma(\omega+\gamma)+(\omega\,p_I+\gamma)\,\hat F_I),$$
and $C_1$ and $C_2$ are 
\begin{align}
\label{eq:C12}
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& (\omega+\gamma)\gamma^2\,\theta\_w+(\theta\_w-\theta\_c)\gamma\,\omega\,p_I\,\hat F_I,
\end{align}
where $\hat F_I$ is given in Eq.\,\eqref{eq:fi}.
Note that for analysis brevity, we let $N_0=1$, thus $S\_u^*$ and $S\_n^*$ are in the scale of 0 to 1.
$\Rnum$ is in the same form as in Eq. \eqref{R0}  
$$\Rnum= \frac{\beta}{\gamma} (1-\Delta),$$
where 
\begin{equation}
\label{eq:del4}
\Delta= \frac{1}{C}\big(C_1\,S\_u^*+((C_1+C_2)(1-\theta\_w)+C\,\theta\_w)\,S\_n^*\big).
\end{equation}

@@

{\bf The second goal} is to explore how changes in \percap testing intensity $\rho$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\rho}$, which specifically depends on $\pder[\Delta]{\rho}$. We use the derived expressions for $S\_u^*$ and $S\_n^*$, given by Eqs.\,\eqref{dfe}, in $\Delta$ \eqref{eq:del4_theta}. Also, we define $\phi = \hat F_S = \frac{\rho \omega}{\omega-\rho}$, to reparametrize $\rho$. This is mainly to avoid singularity in $\hat F_I$ \eqref{eq:fi}, when testing intensity $\rho$ is very close to the rate of test return $\omega$. Thus, $\rho$ is reparametrized as 
\begin{equation}
\label{eq:phi}
\rho=\frac{\omega \phi}{\omega+\phi}.
\end{equation}
This one-to-one monotonic reparametrization enables us to simplify the mathematical expressions and explore the simpler $\pder[\Delta]{\phi}$ instead of the complicated $\pder[\Delta]{\rho}$.
% with an equivalent but simplified $\Delta$ in Eq.\,\eqref{eq:abc} as
% \begin{equation}
% \label{eq:del3}
% \Delta= \frac{K_1\,\phi^2+K_2\,\phi}{K_3},
% \end{equation}
% where
% \begin{align}
% \label{eq:K}
% K_1=& \frac{w_I}{w_S}\Big(((1+p_I)\omega+\gamma)\gamma\theta\_w+\omega^2 p_I\theta\_c \Big), \\
% K_2=& (\omega+\gamma)\Big( (\frac{w_I}{w_S}\omega+\gamma)\gamma\theta\_w + \frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
% K_3=& (\omega+\gamma) \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)(\omega+\phi).
% \end{align}
The derivative is
\begin{align}
\label{eq:dd3dr}
\partial\Delta/\partial\phi=& \frac{1}{d_3} (a_3\,\phi^2+b_3\,\phi+c_3),
\end{align}
where
\begin{multline}
\label{eq:a3}
a_3 = \frac{w_I}{w_S}
\Bigg(\Big(
( \theta_{{w}}{{\it p\_I}}^{2}-\theta_{{c}}{{\it p\_I}}^
{2}\theta_{{w}} ) {\omega}^{3} 
+ ( 2\,\theta_{{w}}{{\it p\_I
}}^{2}-\theta_{{c}}{{\it p\_I}}^{2}-{\it p\_I}\,{\theta_{{w}}}^{2}-
\theta_{{c}}{\it p\_I}\,\theta_{{w}}-{{\it p\_I}}^{2}{\theta_{{w}}}^{2
}+2\,\theta_{{w}}{\it p\_I} ) \gamma\,{\omega}^{2} \\
 + ( -2\,{ \it p\_I}\,{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}+3\,\theta_{{w}}{
\it p\_I}-{\theta_{{w}}}^{2}+\theta_{{w}} ) {\gamma}^{2}\omega 
 + ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {\gamma}^{3} \Big) \frac{w_I}{w_S}
 + ( -\theta_{{c}}{\it p\_I}\,\theta_{{w}}+\theta_{{c}}{\it p\_I}
 ) \gamma\,{\omega}^{2}+ \\
 ( \theta_{{w}}+\theta_{{c}}{\it
p\_I}-{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}\,\theta_{{w}} )
{\gamma}^{2}\omega+ ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {
\gamma}^{3}
\Bigg),
\end{multline}
\begin{align}
\label{eq:abcd2}
b_3&= 2\frac{w_I}{w_S}(\omega+\gamma)\gamma\Big( 
(\omega+\gamma+\omega\,p_I)(2-\theta\_w)\,\gamma\,\theta\_w+ (1-\theta\_w)\,\omega^2 p_I\theta\_c+\omega^2 p_I\theta\_w 
\Big) ,\\
c_3&= (\omega+\gamma)^2\,\gamma\Big( 
(2-\theta\_w)\,\gamma^2\,\theta\_w+
(1+\frac{w_I}{w_S})\,\omega\,\gamma\,\theta\_w+
\frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
d_3&= \frac{(\omega+\gamma)}{\omega} \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)^2(\omega+\phi)^2. 
\end{align}
Note that $\phi\geq 0$, also $b_3$, $c_3$ and $d_3$ are all positive. However $a_3$ can be positive or negative.
If $a_3\geq 0$, $\partial\Delta/\partial\phi \geq 0$ for all feasible range of parameters, thus $\pro\Rnum \leq 0$. 
If $a_3 < 0$, then the quadratic expression in the numerator of \eqref{eq:dd3dr} has a positive root, $\phi^*$, such that for $\phi>\phi^*$, $\partial\Delta/\partial\phi < 0$. An example of this counter-vailing effect of $\rho$ on $\Rnum$ occures when $\theta\_w=0$ and $\theta\_c=1$. This is illustraited in the top-right panel of the Fig.\,\ref{pan2} panel (b), where the strength of isolation for awaiting people is the least, but the most for the confirmed cases. In this case $$a_3=\frac{w_I}{w_S}\,\omega\,\gamma\,p_I((\omega+\gamma)-\frac{w_I}{w_S}(\omega\,p_I+\gamma)).$$
Specifically, in the case of targeted testing which is identified with $\frac{w_I}{w_S}> 1$, and using a perfect sensative test, thus $p_I=1$, there exists a range for $\rho$ over which $\pder[\Rnum]{\rho}\leq\,0$.  
Note that $\rho$ and $\omega$ have a similar mechanism in delaying people to get into $I\_c$, thus we would expect to see the non-trivial counter-vailing effect of these two parameters on $\Rnum$.

>>>>

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where the term $\frac{\beta}{\gamma}$ is the classical basic reproduction number for a SIR model without testing and isolation (see, e.g., \cite{keeling2011modeling}), and $\Delta$ is the reduction parameter defined as follows. 
\begin{equation}
\label{eq:del4}
\Delta= \frac{1}{C}\big(C_1\,S\_u^*+(C_2(1-\theta\_w)+C\,\theta\_w)\,S\_n^*\big),
\end{equation}
where
\begin{align}
\label{eq:C12}
C =& (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big),\\
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& \Big( \omega\gamma(1+p_I)\hat F_I+\gamma^2\,(\omega+\gamma+\hat F_I)\Big)\theta\_w + \omega^2\,p_I\,\hat F_I\,\theta\_c.
\end{align}
Further details of derivation of $\Rnum$ are provided in appendix sec.\,\ref{app:R0}. Further, we show the nontrivial-nonmonotonic effect of testing intensity on $\Rnum$ appendix (See Sec.\,\ref{app:rho}). It is notable that this non-monotonic effect also occures with the rate of test return $\omega$. We show this for the case when testing intesity is very small by using the Taylor approximation of $\Rnum$ at $\rho=0$. \ali{some of these are methods!}

@@

The expression of $\Rnum$ has a complicated form with all of the model parameters involved. This expression can be simplifed and represented given the specific form of matrix $G_{11}$ \eqref{G11}. For the purpose of simplicity we present $\Rnum$ in the manuscript in terms of expressions $C$, $C1$ and $C2$, specified in \eqref{eq:C12}.

@@

This is to provide a mathematical materials to prove that $\pro \Delta$ can be positive or negative, with $\Delta$ defined in Eq.\,\eqref{eq:C12}, and thus $\pro \Rnum < 0$, where $\Rnum$ is the basic reproduction number, given in Eq.\,\eqref{R0}. We can rewrite matrix $G_{11}$ in \eqref{G11} in the following form to simplify the calculations.
\begin{equation}
\label{G112}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c}  S\_u^*/N_0 \\ (1-\theta\_w)\,S\_n^*/N_0  \end {array} \right]
\left[ \begin {array}{cccc} 
C-C_1, C-C_2\end {array} \right],
\end{equation}
where $C$ is the same as the one in Eq. \eqref{eq:C12}, i.e., 
$$C=(\omega+\gamma)(\gamma(\omega+\gamma)+(\omega\,p_I+\gamma)\,\hat F_I),$$
and $C_1$ and $C_2$ are 
\begin{align*}
% \label{eq:C12}
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& \big( \omega\gamma(1+p_I)\hat F_I+\gamma^2\,(\omega+\gamma+\hat F_I)\big)\theta\_w + \omega^2\,p_I\,\hat F_I\,\theta\_c,
\end{align*}
where $\hat F_I$ is given in Eq.\,\eqref{eq:fi}.
Note that for analysis brevity, we let $N_0=1$, thus $S\_u^*$ and $S\_n^*$ are in the scale of 0 to 1.
$\Rnum$ is in the same form as in Eq. \eqref{R0}  
$$\Rnum= \frac{\beta}{\gamma} (1-\Delta),$$
where 
\begin{equation*}
% \label{eq:del4}
\Delta= \frac{1}{C}\big(C_1\,S\_u^*+(C_2(1-\theta\_w)+C\,\theta\_w)\,S\_n^*\big).
\end{equation*}

@@

{\bf The second goal} is to explore how changes in \percap testing intensity $\rho$ affects $\Rnum$. Mathematically we would like to verify the sign of $\pder[\Rnum]{\rho}$, which specifically depends on $\pder[\Delta]{\rho}$. We use the derived expressions for $S\_u^*$ and $S\_n^*$, given by Eqs.\,\eqref{dfe}, in $\Delta$ \eqref{eq:del4_theta}. Also, we define $\phi = \hat F_S = \frac{\rho \omega}{\omega-\rho}$, to reparametrize $\rho$. This is mainly to avoid singularity in $\hat F_I$ \eqref{eq:fi}, when testing intensity $\rho$ is very close to the rate of test return $\omega$. Thus, $\rho$ is reparametrized as 
\begin{equation}
\label{eq:phi}
\rho=\frac{\omega \phi}{\omega+\phi}.
\end{equation}
This one-to-one monotonic reparametrization enables us to simplify the mathematical expressions and explore the simpler $\pder[\Delta]{\phi}$ instead of the complicated $\pder[\Delta]{\rho}$.
The derivative is
\begin{align}
\label{eq:dd3dr}
\partial\Delta/\partial\phi=& \frac{1}{d_3} (a_3\,\phi^2+b_3\,\phi+c_3),
\end{align}
where
\begin{multline}
\label{eq:a3}
a_3 = \frac{w_I}{w_S}
\Bigg(\Big(
( \theta_{{w}}{{\it p\_I}}^{2}-\theta_{{c}}{{\it p\_I}}^
{2}\theta_{{w}} ) {\omega}^{3} 
+ ( 2\,\theta_{{w}}{{\it p\_I
}}^{2}-\theta_{{c}}{{\it p\_I}}^{2}-{\it p\_I}\,{\theta_{{w}}}^{2}-
\theta_{{c}}{\it p\_I}\,\theta_{{w}}-{{\it p\_I}}^{2}{\theta_{{w}}}^{2
}+2\,\theta_{{w}}{\it p\_I} ) \gamma\,{\omega}^{2} \\
 + ( -2\,{ \it p\_I}\,{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}+3\,\theta_{{w}}{
\it p\_I}-{\theta_{{w}}}^{2}+\theta_{{w}} ) {\gamma}^{2}\omega 
 + ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {\gamma}^{3} \Big) \frac{w_I}{w_S}
 + ( -\theta_{{c}}{\it p\_I}\,\theta_{{w}}+\theta_{{c}}{\it p\_I}
 ) \gamma\,{\omega}^{2}+ \\
 ( \theta_{{w}}+\theta_{{c}}{\it
p\_I}-{\theta_{{w}}}^{2}-\theta_{{c}}{\it p\_I}\,\theta_{{w}} )
{\gamma}^{2}\omega+ ( \theta_{{w}}-{\theta_{{w}}}^{2} ) {
\gamma}^{3}
\Bigg),
\end{multline}
\begin{align}
\label{eq:abcd2}
b_3&= 2\frac{w_I}{w_S}(\omega+\gamma)\gamma\Big( 
(\omega+\gamma+\omega\,p_I)(2-\theta\_w)\,\gamma\,\theta\_w+ (1-\theta\_w)\,\omega^2 p_I\theta\_c+\omega^2 p_I\theta\_w 
\Big) ,\\
c_3&= (\omega+\gamma)^2\,\gamma\Big( 
(2-\theta\_w)\,\gamma^2\,\theta\_w+
(1+\frac{w_I}{w_S})\,\omega\,\gamma\,\theta\_w+
\frac{w_I}{w_S} \omega^2 p_I\theta\_c \Big),\\
d_3&= \frac{(\omega+\gamma)}{\omega} \Big((\omega p_I+\gamma)\frac{w_I}{w_S}\phi\,+\,(\omega+\gamma)\gamma \Big)^2(\omega+\phi)^2. 
\end{align}
Note that $\phi\geq 0$, also $b_3$, $c_3$ and $d_3$ are all positive. However $a_3$ can be positive or negative.
If $a_3\geq 0$, $\partial\Delta/\partial\phi \geq 0$ for all feasible range of parameters, thus $\pro\Rnum \leq 0$. 
If $a_3 < 0$, then the quadratic expression in the numerator of \eqref{eq:dd3dr} has a positive root, $\phi^*$, such that for $\phi>\phi^*$, $\partial\Delta/\partial\phi < 0$. An example of this counter-vailing effect of $\rho$ on $\Rnum$ occures when $\theta\_w=0$ and $\theta\_c=1$. This is illustraited in the top-right panel of the Fig.\,\ref{pan2} panel (b), where the strength of isolation for awaiting people is the least, but the most for the confirmed cases. In this case $$a_3=\frac{w_I}{w_S}\,\omega\,\gamma\,p_I((\omega+\gamma)-\frac{w_I}{w_S}(\omega\,p_I+\gamma)).$$
Specifically, in the case of targeted testing which is identified with $\frac{w_I}{w_S}> 1$, and using a perfect sensative test, thus $p_I=1$, there exists a range for $\rho$ over which $\pder[\Rnum]{\rho}\leq\,0$.  
Note that $\rho$ and $\omega$ have a similar mechanism in delaying people to get into $I\_c$, thus we would expect to see the non-trivial counter-vailing effect of these two parameters on $\Rnum$.
